---
title: "Processing in R"
author: "Matt Bihrle, Emily Dailey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Downloading the Package
Ben Harden made a cool package to streamline some of our data processing. to install the package on an internet connected computer call: 

```{r, eval=FALSE}
devtools::install_github("mattbihrle/seaprocess", 
                         build_vignettes = T, 
                         dependencies = T)
```

## Pre-Cruise Setup

1.  Open RStudio. Go to File \> New Project. Then click "New Directory"

2.  Click "Initialize Cruise Project" or "Initialize Cruise Project
    Example", if you want to practice with old data first.

3.  Then, name the project with the cruise name. For example, C303G.

**NOTE:** When you create this project RStudio creates a folder with
your `cruiseID` as the name of the folder and populates this folder with
a bunch of blank templates and other folders. It also creates: \* A
project file with the same name as the `cruiseID` (which will show up in
the file directory as <cruiseID> next to a icon with an R inside a
cube). \* A template processing script (called
`<cruiseID>\_process_data.R` - described below) \* A template eoc
processing script (called \`<cruiseID>\_eoc_process.R\`)

The top left of your screen should read : "<your cruise ID> - RStudio".
This shows you that RStudio is open in project mode and the name of the
project is (<your cruiseID>).

**Note on Projects:**

Projects in RStudio are a way to organize your work so that RStudio
recognizes the current folder as your root directory. It makes it easier
to locate subfolders when you run your code and ultimately makes it much
easier to share this work with other computers/people who might want to
revisit this data. As described above in the note, the project file is
created in the folder the initalization just created. **To make things
run smoothly it's important that you open this project in RStudio when
you reopen RStudio.** To do this, open RStudio by double clicking on the
project file in the folder rather than by double-clicking the RStudio
program icon on th desktop. To load the processing script, you need to
click the the second link in the bottom right hand corner called
"<cruiseID>\_process_data.R". This will load the template processing
script on the left hand side of the screen.

-   In order for Seaprocess to access the folders it needs, you need to
    define some file paths. Next to cruiseID, should be the name of your
    cruise, such as C308D.
-   Where it says `elg_folder <- "<enter-location-of-elg-folder>"`,
    enter the location of the elg folder, which is currently
    `elg_folder <- "Y:/EventData/Event60sec_2021/"`.
-   Where it says `adcp_folder <- "<enter-location-of-adcp-folder>"` ,
    enter the location of the ADCP folder where LTA files are kept which
    is `adcp_folder <- "X:/"`
-   Where it says `ctd_folder <- “<enter-location-of-ctd-folder>"`,
    enter the folder where the cnv files are kept, NOT the average
    folder. Eg: `<- “C:/Data/CTD/Cnv”`
-   Where it says `ros_folder <- “<enter-location-of-ros-folder>"` enter
    the file path the the ros files. This should be the same as the cnv
    folder so you can enter `ros_folder <- ctd_folder`

### Datasheets

The R script will automatically create datasheets for our most common
deployments (ctd, niskin bottles, surface stations, neuston nets, and
meter nets), and autofill the file paths. Double check that the paths
are correct, however you should not need to change anything.

### Adding Datasheets

-   To create a new datasheet, go to RData/[Your cruise]/datasheets.Here
    you should see a "<cruiseID>\_template_input.xls" excel sheet.
    Create a copy of template and rename to match your deployment type
    (C308D_secchi_input.xls, or C308D_sediment_input.xls).
-   When you open your new data sheet you will see a blank sheet with a
    single column for the station number.
-   Click Review \> Unprotect Sheet. This will allow you to add column
    labels to the grey heading cells.
-   R will automatically add surface temp, salinity, chla, station start
    and end, station distance bottom depth and max wire tension. You
    only need to add columns for variables that have to be hand entered!
-   Once you have added your desired column labels be sure to Protect
    Sheet and set the data type on the whole sheet to Text. R is picky
    about how it sees data so may show an error if Excel changed the
    data type. Dates and times are particularly bad.
-   Once you have created your datasheet you need to make sure the
    script can recognize it and add the relevant metadata.
    -   In the "datasheets" portion of the script create a new variable
        with the same name as the datasheet you created and point it to
        the new datasheet.
        -   Eg `secchi_input <- “C:/datasheets/C308E_secchi_input.xls”`
    -   Next scroll down to the "Create Datasheets" portion of the
        script and type the following code:
        -   `create_datasheet(secchi_input, data_type = "SD", cruiseID = cruiseID)`
        -   The "secchi_input" command should match the variable you
            just created above and the `data_type` should be a two or
            three letter code that matches the code you will input in
            the summary datasheet.

## Data Entry

### Station Summary Sheet

-   Data entry will be less time consuming than our previous method but
    will look a little different. The most important sheet to keep up to
    date is the `summary_input`. In this sheet, enter the station number
    formatted as "C303F-001", the deployment type, the date
    (yyyy-mm-dd), time of start(eg. 15:32), time of end for neuestons
    (and anything you want max tension recorded), and ZD.

    -   You will notice that there are a few darker grey cells. They can
        be left blank unless you want to override the automatic metadata
        that the script adds to the output file.

### Deployment Specific Datasheet

-   On the deployment datasheet you only need to fill in the station
    number (formatted as C303F-001) and the data that needs to be hand
    entered. **NOTE: If a column is made up of numbers, only input
    numbers.** Characters like "\<", "\>" "\~" will result in NA values.

### Hydrocast Data Entry

\*For hydrocasts, two datasheets will need to be filled out. - First,
the ctd_input will contain all of the meta data for the cast - Second,
bottle_input will need station number and ONLY the bottles that you drew
water from. This is a change from our previous practice in hydrowork of
copying and pasting data from all 12 bottles.

### Surface Stations

Surface stations are treated as "bottles" and their data is entered into
the bottle_datasheet. For their metadata there are three options:

#### As Bottle 13

-   For this option, enter their bottle number as 13 and the HC metadata
    will be associated with that bottle

#### As Own Deployment

-   This option requires the Surface station entered in the
    summary_input sheet as `SS.` It's metadata will be it's own but it
    will be grouped with the station. Still enter it's bottle data in
    `bottle_input` and record the bottle number as `SS`.

#### As Own Station

-   The easiest way to count surface stations separately from all other
    deployments is to start counting at 100 (SS-001 will be SS-101,
    SS-002 will be SS-102 etc.). Still input the data into
    `summary_input` and `bottle_input` as `SS`

## Running the Script

Once everything is setup, click on the top of the script and press
ctrl + enter to individually run each line of code. R may show a
"warning" on some lines. These are okay. If you see an "error," double
check that you entered the relevant file paths correctly, you have no
typos or stray characters and that your slashes are all forward ("/"
instead of ""). To run the whole script at once you can go to: Code\>Run
Region\>Run all or hit ctrl + alt + A. If you are running the whole
script make sure to comment out any lines you are not using by putting
a"\#" at the beginning of the line. If R see a blank data sheet it will
show and error and stop the run where the error occurred. In the
"output" folder R will create a csv file (comma separated value) for
each datasheets as well as a text file formatted for ODV input with the
"output/odv" folder. As of July 2023, the ODV text files are only for
our most common deployments however you can always import csv files into
ODV.

## Upon EOC

The `eoc_process` script will lead you through the necessary steps to copy the needed files over to the raw folders. Other than running through that script, the end of cruise second sci work will look the same. 

## Tips and Tricks

-   Keep in mind that for these files to be found, the ADCP and
    Datalogger computers need to be on, as R pulls directly from those
    computers.
-   If an error says that there is no input folder, make sure your
    cruise ID is correct, that the computer is on, and that there is in
    fact an elg lta file, or station summary to pull from
-   Formatting is important! If the script is showing errors, the first
    thing to check is whether you have any formatting inconsistencies in
    your datasheets.
-   SeaProcess will overwrite the outputs each time it is run, this
    means that if you mess anything up in the outputs folder, just run
    "Process_data" again to refresh your files.
-   The `process_data` script is designed to be simple so if something
    is changed by accident it should be easy enough to go to a previous
    cruise, copy and paste their script and update it for your own
    cruise.
-   For more information on what is happening in the background call:
```{r, eval = FALSE}
library(seaprocess)
vignette("seaprocess")
```

-   For more information on variables included in the packages call:
```{r, eval = FALSE}
library(seaprocess)
vignette("variables")
```
