---
title: "Setup and EOC SeaProcess"
author: "Matt Bihrle, Emily Dailey"
date: "2025-Jan-14"
output:
  rmarkdown::html_vignette: 
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Setup and EOC SeaProcess}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**NOTE:** If SeaProcess is not installed already see the [Package
Management for SeaProcess](Package_Management_for_seaProcess.html)
vignette or documentation in RCS Science or CC Science.

# Pre-Cruise Setup

1.  Open RStudio. Go to File \> New Project. Then click "New Directory"

2.  Click "Initialize Cruise Project" or "Initialize Cruise Project
    Example", if you want to practice with old data first.

3.  Then, name the project with the cruise name. For example, C303G.

**NOTE:** When you create this project, RStudio creates a folder with
your `cruiseID` as the name of the folder and populates this folder with
a bunch of blank templates and other folders. It also creates:

-   A project file with the same name as the `cruiseID` (which will show
    up in the file directory as <cruiseID> next to a icon with an R
    inside a cube).

-   A template processing script (called `<cruiseID>_process_data.R` -
    described below)

-   A template eoc processing script (called `<cruiseID>_process_eoc.R`)

The top left of your screen should read : "<your cruise ID> - RStudio".
This shows you that RStudio is open in project mode and the name of the
project is (<your cruiseID>).

**To make things run smoothly it's important that you open this project
in RStudio when you reopen RStudio.** To do this, open RStudio by double
clicking on the project file in the folder rather than by
double-clicking the RStudio program icon on the desktop. To load the
processing script, you need to click the the second link in the bottom
right hand corner called `<cruiseID>_process_data.R`. This will load the
template processing script on the left hand side of the screen.

**Note on Projects:** Projects in RStudio are a way to organize your
work so that RStudio recognizes the current folder as your root
directory. It makes it easier to locate subfolders when you run your
code and ultimately makes it much easier to share this work with other
computers/people who might want to revisit this data. As described above
in the note, the project file is created in the folder the initalization
just created.

-   In order for Seaprocess to access the folders it needs, you need to
    define some file paths. Next to cruiseID, should be the name of your
    cruise, such as C308D.
-   Where it says `elg_folder <- "<enter-location-of-elg-folder>"`,
    enter the location of the elg folder on the datalogger computer. For
    example, on Cramer: `elg_folder <- "Y:/EventData/Event60sec_2021/"`.
-   Where it says `adcp_folder <- "<enter-location-of-adcp-folder>"` ,
    enter the location of the ADCP folder where LTA files are kept. Eg.
    `adcp_folder <- "X:/"`
-   Where it says `ctd_folder <- “<enter-location-of-ctd-folder>"`,
    enter the folder where the .avg files are kept. Eg:
    `<- “C:/Data/CTD/Avg”`
-   Where it says `ros_folder <- “<enter-location-of-ros-folder>"` enter
    the file path the the ros files. This should be the same as the cnv
    folder so you can enter`ros_folder <- "C:/Data/CTD/Cnv`

## Datasheets

The R script will automatically create datasheets for our most common
deployments (ctd, niskin bottles, surface stations, neuston nets, meter
nets, and secchi discs), and autofills the file paths. Double check that
the paths are correct, however you should not need to change anything.

## Adding Datasheets

-   To create a new datasheet, go to RData/[Your cruise]/datasheets.
    Here you should see a "<cruiseID>\_template_input.xls" excel sheet.
    Create a copy of template and rename to match your deployment type
    (C308D_secchi_input.xls, or C308D_sediment_input.xls).
-   When you open your new data sheet you will see a blank sheet with a
    single column for the station number.
-   Click Review \> Unprotect Sheet. This will allow you to add column
    labels to the grey heading cells.
-   R will automatically add surface temp, salinity, chla, station start
    and end, station distance bottom depth and max wire tension. You
    only need to add columns for additional variables that have to be
    hand entered!
-   Once you have added your desired column labels be sure to "Protect
    Sheet" and set the data type on the entire sheet to "Text." R is
    picky about how it sees data and may show an error if Excel changed
    the data type. Dates and times are particularly bad.
-   Once you have created your datasheet you need to make sure the
    script can recognize it and add the relevant metadata.
    -   In the "datasheets" portion of the script create a new variable
        with the same name as the datasheet you created and point it to
        the new datasheet.
        -   Eg `secchi_input <- “datasheets/C308E_secchi_input.xls”`
    -   Next scroll down to the "Create Datasheets" portion of the
        script and type the following code:
        -   `create_datasheet(secchi_input, data_type = "SD", cruiseID = cruiseID)`
        -   The "secchi_input" command should match the variable you
            just created above and the `data_type` should be a two or
            three letter code that matches the code you will input in
            the summary datasheet.

**Pro-Tip:** The `create_summary` function will work to fill metadata
for any datasheet that has at least the same columns as the autocreated
"summary_input" sheet.This means we can have a single datasheet for
observational data or surface station data instead of separate summary
sheets for each. To do this for an obs sheet: - Make a copy of the
"summary_input.xls", and rename it "<cruiseID>\_obs_input.xls." - Keep
the columns already present and add the additional columns you want. -
In the `process_data.R` script first define the file path:
`obs_input <- "datasheets/"<cruiseID>_obs_input.xls"` - The create a new
line with `create_summary()` instead of `create_datasheet()`:
`create_summary(summary_input = obs_input, elg_folder, cruiseID = cruiseID, csv_filename = "obs_datasheet.csv")`

## Calculation Sheets

Calculation sheets for water chemistry can be setup to autofill values
into the bottle datasheet.

To do this, ensure there is a folder in your cruise project called
"calc_sheets" (this should be autocreated when you create a new cruise).

Next, double check that your `<cruise_ID>process_data.R` file has the
line `calc_folder <- "calc_sheets"`

Then, add two parameters to your `create_datasheet()` for bottle data :
`process_calc = T` and `calc_folder = calc_folder`. Your bottle
datasheet function should look like:
`create_datasheet(bottle_input, data_type = "bottle", ros_input = ros_folder, cruiseID = cruiseID, process_calc = T, calc_folder = calc_folder)`

Lastly, add your calc sheets! R will look for a tab in your excel file
labeled exactly "output" which has 3 columns: station number, bottle,
and parameter (ph, chla, no3, etc). All the other tabs in the excel file
can contain whatever, so it is easy to reuse our normal SEA calc sheets.

### Adding New Parameters

The bottle datasheet autopopulates with a few typical parameters for SEA
cruises but you can always add your own! Just be sure to add them
somewhere between the pre-existing parameters--not at the beginning or
end-- and ensure they have the correct column names. R will
automatically replaces spaces with `_` and remove `%`, `\#`,periods,
dashes, and parentheses. It is best practice to stick with numbers,
lowercase letters, and `_` for naming columns. *If you are seeing an
error, check your naming in both the bottle datasheet and calculation
sheet!*

The typical SEA parameters (no3, po4, pH, alk, chla) will all auto-fill
with units in the output sheet. If you need units on new parameters be
sure to add them to your column name.

If you don't wish to use the calc sheets function, just take the
`process_calc` and `calc_folder` parameters out of your bottle function
and hand enter data as usual. Any data hand entered will always
overwrite anything auto-filled from the calc sheets so there is a backup
if something is not working.

# End Of Cruise

The `eoc_process.R` script will lead you through the necessary steps to
copy the needed files over to the "raw" folder. As of Jan-2025 this is
the elg files, averaged cnvs and ros files, and LTAs. It will also
create a wire log output for submission to UNOLS. **Please read and
follow all steps.**

The primary goal of this script is setup your cruise's RData folder to
contain all of the raw data and scripts used to create outputs on your
trip. This will allow future scientists and students to access most of
what they need from a cruise through a single folder. It is still
imperative that all data files created on the trip go back to the office
in the places they historically live (see the second scientist checklist
for more info). However, things like .hex files or additional ADCP files
do not need to be stored in the "raw" folder.

# Tips and Tricks

-   Each function within the package has lots of information to
    customize outputs. For more information on a specific function just
    put a `?` infront of it. For example: `?create_summary()`.

-   Keep in mind that for these files to be found, the ADCP and
    Datalogger computers need to be on, as R pulls directly from those
    computers.

-   If an error says that there is no input folder, make sure your
    cruise ID is correct, that the computer is on, and that there is in
    fact an elg, lta file, or station summary to pull from.

-   Formatting is important! If the script is showing errors, the first
    and second thing to check is whether you have any formatting
    inconsistencies in your datasheets.

-   SeaProcess will overwrite the outputs each time it is run, this
    means that if you mess anything up in the outputs folder, just run
    `process_data` again to refresh your files.

-   The `process_data` script is designed to be simple so if something
    is changed by accident it should be easy enough to go to a previous
    cruise, copy and paste their script and update it for your own
    cruise.

-   `create_summary` has an option to autofill max tension from either
    the event file created by SCS or the .RAW file created by SCS (RCS
    only).

    -   To process wire tension through the raw file define where the
        .raw file is found (`lci_raw_folder <- "W://"`) and add the
        following parameters to your create summary function:
        `raw_folder = lci_raw_folder` and `process_lci = T`
    -   Processing the second to second data will take longer than
        minute to minute so be patient as the program runs! See
        `?create_summary` for more information.

-   `process_elg` has a couple options to filter out bad data.

    -   It will automatically filter out all flow_through data when
        salinity drops below a a threshold. This can be set by the user
        by adding the argument `min_sal = <desired sal threshold>`. Call
        `?filter_elg` for more information and to see the current
        default settings.
    -   Additionally, there is an option to set custom filters for all
        elg variables. If this is desired, your `process_elg` function
        will look something like:

    ```{r eval = FALSE}

    process_elg(elg_folder, cruiseID = cruiseID, custom_filter = TRUE, 
        filter_params =c("max_fluor = 30", "min_temp = 15, max_wind_sp = 66"))
    ```

    Again for more information on formatting, call `?filter_elg`.

-   For more information on what is happening in the background of the
    package call:

```{r, eval = FALSE}
library(seaprocess)
vignette("introduction_to_seaprocess")
```

or click here: [Introduction TO
SeaProcess](introduction_to_seaprocess.html)

-   For more information on variables included in the packages call:

```{r, eval = FALSE}
library(seaprocess)
vignette("variables")
```

or click here: [Variables](variables.html)
