---
title: "variables"
author: "Matt Bihrle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(seaprocess)
library(knitr)
library(kableExtra)
metadata <- readr::read_delim("metadata/metadata.txt")
```

# SeaProcess Variables

Overview of all variables used in the SeaProcess package organized by deployment type. Just run the code chunk below your desired deployment to generate a table. A master table is located at the end of the document.

## Neuston Tow

```{r, eval=TRUE}
neuston <- dplyr::filter(metadata, (stringr::str_detect(deployment, "NT") | deployment == "all"))
kable_material(kable(neuston, col.names = stringr::str_to_title(names(neuston))), fixed_thead = T)
```

For more information on calculations related to neuston tows, see: `?seaprocess::compile_neuston`

## Meter Net

```{r, eval=TRUE}
meter <- dplyr::filter(metadata, (stringr::str_detect(deployment, "MN") | deployment == "all"))
kable_material(kable(meter, col.names = stringr::str_to_title(names(meter))), fixed_thead = T)
```

For more information on calculations related to meter nets and other nets deployed on the wire, see: `?seaprocess::compile_meter` 

## Hydrocast

```{r, eval=TRUE}
hydro <- dplyr::filter(metadata, (stringr::str_detect(deployment, "HC") |  stringr::str_detect(deployment, "CTD") | deployment == "all"))
kable_material(kable(hydro, col.names = stringr::str_to_title(names(hydro))), fixed_thead = T)
```

For more information on hydrocast data processing see `?seaprocess::compile_bottle` <!-- MB TODO, expand this section to point towards info on ctd processing as well -->

## ADCP

```{r, eval=TRUE}
adcp <- dplyr::filter(metadata, (stringr::str_detect(deployment, "ADCP") | deployment == "all"))
kable_material(kable(adcp, col.names = stringr::str_to_title(names(adcp))), fixed_thead = T)
```

For more information on ADCP processing see: `?seaprocess::read_adcp`, `?seaprocess::calc_echo_amp`

## Hourly

```{r, eval=TRUE}
hourly <- dplyr::filter(metadata, (stringr::str_detect(deployment, "hourly") | deployment == "all"))
kable_material(kable(hourly, col.names = stringr::str_to_title(names(hourly))), fixed_thead = T)
```

For more information on hourly/elg processing see `?seaprocess::process_elg`

### Master Variables Table

```{r, eval=TRUE}
kable_material(kable(metadata, col.names = stringr::str_to_title(names(metadata))), fixed_thead = T)
```
